---
title: "ConsensusMetaDA: R package for microbiome analysis using multiple algorithm"
author: "Manoharan Kumar, Matt Field"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
---

# Introduction
This manual demonstrates R analysis techniques.

BiocManager::install("phyloseq")
BiocManager::install("DESeq2")
BiocManager::install("ALDEx2")
BiocManager::install("edgeR")
BiocManager::install("ADAPT")
BiocManager::install("edgeR")
BiocManager::install("metagenomeSeq")

# consensusDE examples

Begin by first installing and loading the ```consensusDE``` library. To illustrate functionality of ```consensusDE```, we will utilise RNA-seq data from the ```airway``` and annotation libraries as follows. Begin by installing and attaching data from these libraries as follows:

## Load Libraries

```{r echo=TRUE, message=FALSE, warning=FALSE}
# load consensusDE
library(ConsensusMetaDA)


# Load the packages
library(ggplot2)
library(phyloseq)
library(DESeq2)
library(ALDEx2)
library(edgeR)
library(ADAPT)
library(metagenomeSeq)

```

## Volcano

This plot a volcano plot, which compares the Log-fold change versus significance of change -log transformed score. As above and described in the MA plot section, to use independently of ```OTUs_multi_DE()``` and plot to only one comparison, constructing a list with one data.frame with the columns labelled "ID", "AveExpr", and "Adj_PVal" is required.

```{r echo=TRUE, message=FALSE}
OTU_plots(merged_in = comparison_list,
           name = "untrt-trt",
           volcano = TRUE)
```

