---
title: 'ConsensusMetaDA: an R Package for consensus-based differential abundance analysis of metagenomic data.'
author: "Manoharan Kumar, Matt Field"
date: 'Last modified: 2026-09-01. Compiled: `r Sys.Date()`'
output:
  html_document:
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
vignette: >
  %\VignetteIndexEntry{ConsensusMetaDA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction to ConsensusMetaDA

ConsensusMetaDA is an R package for microbiome differential abundance (DA)
analysis using multiple algorithms and reaching a consensus. No single tool
performs best on all metagenomic datasets, so a consensus approach yields more
robust differentially abundant taxa. ConsensusMetaDA runs **seven** widely used
tools — edgeR, DESeq2, ALDEx2, metagenomeSeq, ADAPT, ANCOM-BC2, and MaAsLin3 —
takes standard biom + sample table input, and also produces microbiome
visualisations (alpha diversity, beta diversity, rarefaction, bidirectional and
scaled plots).

Consensus is reached three ways: vote rules (union / k-of-7 / intersect),
p-value combination (Fisher, Stouffer, and the dependence-robust Cauchy/ACAT),
and rank-sum aggregation.

ConsensusMetaDA is organised around two core functions:

* ```build_OTU_counts()``` — build (or filter) a phyloseq object from a biom
  file + sample table (+ optional taxonomy), or from an existing phyloseq object.
* ```OTUs_multi_DA()``` — perform DA analysis across all pairwise comparisons.

Plotting is provided by ```OTUs_plots()```.

```{r install, eval=FALSE}
# ADAPT requires R >= 4.4 (use R 4.4.2)
# if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# for (pkg in c("phyloseq","DESeq2","ALDEx2","edgeR","metagenomeSeq","ANCOMBC"))
#   if (!require(pkg, character.only = TRUE)) BiocManager::install(pkg)
# # ADAPT, maaslin3 as per their install instructions
# if (!require("ConsensusMetaDA", quietly = TRUE)) {
#   if (!require("remotes", quietly = TRUE)) install.packages("remotes")
#   remotes::install_github("kmanoharan01/ConsensusMetaDA")
# }
```

## Load libraries

```{r libraries}
library(ggplot2)
library(phyloseq)
library(vegan)
library(ConsensusMetaDA)
```

# Example data

We use the bundled `ecc_saliva` dataset (early-childhood-caries saliva
microbiome). It is available both as example files under `inst/extdata/` and as
a ready-made phyloseq object via `data()`.

```{r data}
library(ADAPT)
data("ecc_saliva")
ecc_saliva

# resolve bundled example files (works regardless of working directory)
ecc_saliva_biom         <- system.file("extdata", "ecc_saliva_Otu.biom",
                                        package = "ConsensusMetaDA")
ecc_saliva_sample_table <- system.file("extdata", "ecc_saliva_sample_table.txt",
                                        package = "ConsensusMetaDA")

# taxonomy taken from the bundled object so it matches the biom's ASV ids
ecc_saliva_tax <- tax_table(ecc_saliva)
```

# Building a phyloseq object

`build_OTU_counts()` has two entry points and optional filters
(`abundance_threshold`, `prevalence_threshold`, `rarity_threshold`,
`variance_threshold`). **Because the first argument is now `physeq`, always call
it with named arguments.**

## From files (biom + sample table + taxonomy)

```{r build-files}
ecc_saliva_build_OTU_counts <- build_OTU_counts(
  biom                 = ecc_saliva_biom,
  sample_table         = ecc_saliva_sample_table,
  tax_tables           = ecc_saliva_tax,
  abundance_threshold  = 10,
  prevalence_threshold = 0.1,
  rarity_threshold     = 0,
  variance_threshold   = 0
)
ecc_saliva_build_OTU_counts
```

## From a prebuilt phyloseq object

When you already have a phyloseq object, pass it via `physeq =` and use the
function purely for filtering. This is orientation-safe (it transposes
internally when taxa are stored as columns).

```{r build-physeq}
ecc_ps <- build_OTU_counts(physeq = ecc_saliva, prevalence_threshold = 0.1)
ecc_ps
```

## Individual filters

Each filter can be used on its own.

```{r filter-rarity}
build_OTU_counts(biom = ecc_saliva_biom, sample_table = ecc_saliva_sample_table,
                 tax_tables = ecc_saliva_tax, rarity_threshold = 0.1)
```

```{r filter-abundance}
build_OTU_counts(biom = ecc_saliva_biom, sample_table = ecc_saliva_sample_table,
                 tax_tables = ecc_saliva_tax, abundance_threshold = 10)
```

```{r filter-prevalence}
build_OTU_counts(biom = ecc_saliva_biom, sample_table = ecc_saliva_sample_table,
                 tax_tables = ecc_saliva_tax, prevalence_threshold = 0.1)
```

```{r filter-variance}
build_OTU_counts(biom = ecc_saliva_biom, sample_table = ecc_saliva_sample_table,
                 tax_tables = ecc_saliva_tax, variance_threshold = 0.2)
```

# Performing differential abundance

`OTUs_multi_DA()` runs the seven tools on every pairwise comparison of the
grouping variable (default `group_var = "Group"`) and consolidates the results.
An UpSet plot is written per comparison.

```{r da}
ecc_saliva_DA <- OTUs_multi_DA(ecc_saliva_build_OTU_counts,
                               group_var = "Group", fdr_cut = 0.05)

names(ecc_saliva_DA)                     # the contrasts that were run
cons <- ecc_saliva_DA[[1]]$consolidated  # consolidated table for first contrast
```

## The consolidated table

For each taxon the consolidated table carries, per tool, a log fold-change,
p-value, adjusted p-value and significance flag, plus:

* ```n_sig``` — how many of the seven tools called the taxon significant
* ```n_tools``` — how many tools ran
* **combined p-values** (appended automatically):
    * ```p_fisher```, ```padj_fisher```, ```sig_fisher```
    * ```p_stouffer```, ```padj_stouffer```, ```sig_stouffer```
    * ```p_cauchy```, ```padj_cauchy```, ```sig_cauchy```

Because the seven tools share the same count matrix and overlapping models
(i.e. their p-values are correlated), the **Cauchy (ACAT)** combiner is
recommended — it is valid under arbitrary dependence. Fisher and Stouffer assume
independence and are provided as comparators.

```{r combined-cols}
grep("fisher|stouffer|cauchy", names(cons), value = TRUE)
```

There is deliberately **no composite fold-change column**: the tools report
fold-change on different footings (ALDEx2's `effect` is a standardized effect
size, not a log2 fold-change; ANCOM-BC2 / MaAsLin3 / metaSeq each carry their
own normalization), so averaging them would mix different quantities. Read each
tool's `logFC_<tool>` column directly if needed.

## Consensus decision rules

`compare_decision_rules()` puts every strategy side by side — vote rules, the
three combiners, and rank-sum aggregation.

```{r decision-rules}
compare_decision_rules(cons, fdr_cut = 0.05)
```

## Compact summary table

`build_summary_table()` gives per-tool adjusted p-values and ranks, the
rank-sum, and the order-statistic p-values `p_union` / `p_atleast2` /
`p_intersect_all`. Threshold `p_union` for the union, `p_intersect_all` for the
intersect, and use `p_atleast_k()` for any *k*.

```{r summary-table}
st <- build_summary_table(cons)
st$p_atleast5 <- p_atleast_k(st, k = 5)   # "at least 5 of 7 tools" p-value
head(st)

describe_summary_table()                  # column dictionary
```

The summary-table columns are:

* ```<tool>_adj_p``` — BH-adjusted p-value from each tool
* ```<tool>_rank``` — rank of that tool's p-value (smallest = most significant)
* ```rank_sum``` — sum of the per-tool ranks (table is sorted by this)
* ```p_union``` — smallest adjusted p-value across tools (threshold → union)
* ```p_atleast2``` — 2nd-smallest (threshold → "≥2 tools"); `p_atleast_k()` for other k
* ```p_intersect_all``` — largest adjusted p-value (threshold → intersect)

# Plotting functions

`OTUs_plots()` creates five plots — (1) alpha diversity (Shannon), (2) beta
diversity (PCoA), (3) rarefaction curve, (4) scaled taxa-level plot and
(5) bidirectional plots — and saves them as PDFs. It also returns them as
objects for inline display. See `?OTUs_plots`.

```{r plots-run}
ecc_saliva_plots <- OTUs_plots(ecc_saliva_build_OTU_counts)
```

## Rarefaction curve

A rarefaction curve shows how many species you detect as sequencing depth
increases, allowing fair comparison of diversity across samples sequenced to
different depths.

## Alpha diversity

Alpha diversity (Shannon index) measures within-sample richness and evenness.

```{r plot-alpha}
print(ecc_saliva_plots$AlphaDiv$tax_level)
```

## Beta diversity / PCoA

Beta diversity (principal coordinates analysis) shows how microbiome composition
differs among groups and samples.

```{r plot-beta}
print(ecc_saliva_plots$BetaDiv)
```

## Bidirectional plot

A bidirectional plot shows two groups' mean abundance and prevalence together,
richer than either alone.

```{r plot-bidirectional}
print(ecc_saliva_plots$all_comparisons_results$Control_vs_Case$Genus_plot)
```

Interpretation:

* High abundance + high prevalence — beneficial bacteria expanding
* Higher abundance + lower prevalence — pathogen blooming in some individuals
* Lower abundance + higher prevalence — opportunists colonising stressed communities
* Lower abundance + lower prevalence — sensitive bacteria dying off

# Citing ConsensusMetaDA

When using this package, please cite ConsensusMetaDA and every method used.

```{r citation, eval=FALSE}
citation("ConsensusMetaDA")
```

**DA tools**

* **DESeq2** — Love, M.I., Huber, W., Anders, S. (2014). Moderated estimation of
  fold change and dispersion for RNA-seq data with DESeq2. *Genome Biology*
  15(12):550.
* **edgeR** — Robinson MD, McCarthy DJ, Smyth GK (2010). *Bioinformatics* 26,
  139–140. McCarthy DJ, Chen Y, Smyth GK (2012). *Nucleic Acids Research* 40,
  4288–4297.
* **ALDEx2** — Fernandes, A.D. et al. (2013). ANOVA-Like Differential Expression
  (ALDEx) Analysis for Mixed Population RNA-Seq. *PLOS ONE* 8(7), e67019.
* **ADAPT** — Wang, M., Fontaine, S., Jiang, H., Li, G. (2024). ADAPT: Analysis
  of Microbiome Differential Abundance by Pooling Tobit Models. *Bioinformatics*
  40(11).
* **metagenomeSeq** — Paulson JN, Stine OC, Bravo HC, Pop M (2013). Differential
  abundance analysis for microbial marker-gene surveys. *Nature Methods* 10,
  1200–1202.
* **ANCOM-BC2** — Lin, H., Peddada, S.D. (2020). Analysis of compositions of
  microbiomes with bias correction. *Nature Communications* 11, 3514.
* **MaAsLin** — Mallick, H. et al. (2021). Multivariable association discovery in
  population-scale meta-omics studies. *PLoS Computational Biology* 17(11),
  e1009442.

**Consensus / combination methods**

* **Cauchy combination (recommended)** — Liu, Y., Xie, J. (2020). Cauchy
  combination test: a powerful test with analytic p-value calculation under
  arbitrary dependency structures. *JASA* 115(529), 393–402.
* **ACAT** — Liu, Y. et al. (2019). ACAT: a fast and powerful p-value combination
  method for rare-variant analysis. *AJHG* 104(3), 410–421.
* **Fisher's method** — Fisher, R.A. (1932). *Statistical Methods for Research
  Workers*, 4th ed. Oliver and Boyd.
* **Stouffer's Z** — Stouffer, S.A. et al. (1949). *The American Soldier, Vol. 1*.
  Princeton University Press.
* **Rank aggregation** — Guo, Zhao, Ye, Sheng, Shyr (2014).

**Framework**

* **phyloseq** — McMurdie, P.J., Holmes, S. (2013). phyloseq: an R package for
  reproducible interactive analysis and graphics of microbiome census data.
  *PLOS ONE* 8(4), e61217.
```
